---
title: "Analysis of human pancreas data (Baron et al. 2016)"
author: "Michał Kabza"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: default
    highlight: pygments
    df_print: kable
    toc: true
    toc_depth: 2
    toc_float: true
---

# Preparing the environment

Load the required packages:
```{r, message = FALSE}
library(tidyverse)
library(glue)
library(knitr)
library(Matrix)
library(scRNAseq)
library(scater)
library(scran)
library(PCAtools)
library(dittoSeq)
library(Nebulosa)
library(pheatmap)
library(viridis)
```

# Preparing the data

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Load the Baron et al. (2016) human pancreas dataset
sce <- BaronPancreasData('human')
sce
# Cell metadata
colData(sce)
# Cell type summary
sce$label %>%
	table() %>%
	enframe() %>%
	arrange(desc(value))
# Remove cell types with fewer than 100 cells
cell_types_to_keep <- sce$label %>%
	table() %>%
	enframe() %>%
	filter(value >= 100) %>%
	pull(name)
sce <- sce[, sce$label %in% cell_types_to_keep]
sce
# Order cell types by freqency
sce$label <- fct_infreq(sce$label)
colData(sce)
```

# Data normalization

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Solution 1: library sizes (i.e. total sum of counts per cell)
sce <- computeLibraryFactors(sce)
summary(sizeFactors(sce))
# Solution 2: deconvolution strategy implemented in scran
quick_clusters <- quickCluster(sce)
table(quick_clusters)
sce <- computeSumFactors(sce, clusters = quick_clusters)
summary(sizeFactors(sce))
# Calculate log-transformed normalized counts
sce <- logNormCounts(sce)
sce
```

# Selecting highly variable genes

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Model gene variance
dec <- modelGeneVar(sce)
dec
dec %>%
	as.data.frame() %>%
	ggplot(mapping = aes(x = mean, y = total)) +
	geom_point() +
	geom_smooth() +
	labs(
		title = "Modelling gene variance",
		x = "Mean log-expression",
		y = "Total variance"
	) +
	theme_bw()
# Select the top 2000 variable genes
top_hvgs <- getTopHVGs(dec, n = 2000)
length(top_hvgs)
# Select the top 10% of variable genes
top_hvgs <- getTopHVGs(dec, prop = 0.1)
length(top_hvgs)
# Select all variable genes with FDR below 5%
top_hvgs <- getTopHVGs(dec, fdr.threshold = 0.05)
length(top_hvgs)
```

# Principal component analysis

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Set the RNG seed for reproducible results
set.seed(42)
# Solution 1: run the PCA and use scree plot to find the number of significant PCs
sce <- runPCA(sce, ncomponents = 50, subset_row = top_hvgs)
sce
percent_var <- attr(reducedDim(sce, "PCA"), "percentVar")
pca_elbow <- findElbowPoint(percent_var)
pca_elbow
ggplot(mapping = aes(x = seq_along(percent_var), y = percent_var)) +
	geom_point() +
	geom_vline(xintercept = pca_elbow, col = "blue") +
	labs(
		title = "Scree plot",
		x = "PC",
		y = "Variance explained (%)"
	) +
	theme_bw()
reducedDim(sce, "PCA") <- reducedDim(sce, "PCA")[, 1:pca_elbow, drop = FALSE]
ncol(reducedDim(sce, "PCA"))
# Solution 2: run the PCA and remove PCs corresponding to technical noise
sce <- denoisePCA(sce, dec, subset.row = top_hvgs, min.rank = 5, max.rank = 50)
ncol(reducedDim(sce, "PCA"))
```

# Dimensionality reduction by UMAP

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Set the RNG seed for reproducible results
set.seed(42)
# Run UMAP
sce <- runUMAP(sce, dimred = "PCA")
# UMAP plot colored by cell type
dittoDimPlot(sce, "label", reduction.use = "UMAP", size = 0.5,
             main = "UMAP plot colored by cell type", legend.show = FALSE,
             do.label = TRUE, labels.highlight = FALSE, labels.size = 4)
```

# Marker gene detection

## Detecting marker genes

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
markers <- findMarkers(sce, groups = sce$label, test.type = "t",
                       pval.type = "all", direction = "up", lfc = 1)
markers
head(markers[["alpha"]])
```

## Top marker for each cell type

```{r, message = FALSE, warning = FALSE}
top_markers <- lapply(markers, function(x) rownames(x)[1]) %>%
	unlist() %>%
	enframe()
top_markers
```

## Top marker gene tables {.tabset}

```{r, message = FALSE, warning = FALSE, results = 'asis'}
for (cell_type in levels(sce$label)) {
	cat("\n\n###", cell_type, "\n")
	markers[[cell_type]] %>%
		as.data.frame() %>%
		head(n = 5) %>%
		t() %>%
		kable() %>%
		print()
}
```

## Top marker gene expression plots {.tabset}

```{r, message = FALSE, warning = FALSE, results = 'asis'}
top_markers <- lapply(markers, function(x) rownames(x)[1])
for (cell_type in levels(sce$label)) {
	cat("\n\n###", cell_type, "\n")
	expression_plot <- dittoDimPlot(sce, top_markers[[cell_type]], reduction.use = "UMAP",
	                                size = 0.5, order = "increasing")
	print(expression_plot)
}
```

## Top marker gene density plots {.tabset}

```{r, message = FALSE, warning = FALSE, results = 'asis'}
top_markers <- lapply(markers, function(x) rownames(x)[1])
for (cell_type in levels(sce$label)) {
	cat("\n\n###", cell_type, "\n")
	density_plot <- plot_density(sce, top_markers[[cell_type]], size = 0.5)
	print(density_plot)
}
```

## Top marker gene violin plots {.tabset}

```{r, message = FALSE, warning = FALSE, results = 'asis'}
top_markers <- lapply(markers, function(x) rownames(x)[1])
for (cell_type in levels(sce$label)) {
	cat("\n\n###", cell_type, "\n")
	violin_plot <- dittoPlot(sce, top_markers[[cell_type]], group.by = "label",
	                         plots = c("vlnplot"), vlnplot.lineweight = 0.5,
	                         legend.show = FALSE)
	print(violin_plot)
}
```

## Top marker gene ridge plots {.tabset}

```{r, message = FALSE, warning = FALSE, results = 'asis'}
top_markers <- lapply(markers, function(x) rownames(x)[1])
for (cell_type in levels(sce$label)) {
	cat("\n\n###", cell_type, "\n")
	ridge_plot <- dittoPlot(sce, top_markers[[cell_type]], group.by = "label",
	                        plots = c("ridgeplot"), ridgeplot.lineweight = 0.5,
	                        legend.show = FALSE)
	print(ridge_plot)
}
```

## Top marker gene heatmaps

```{r, message = FALSE, warning = FALSE}
top_markers <- sapply(markers, function(x) rownames(x)[1])
# Average gene expression heatmap
plotGroupedHeatmap(sce, top_markers, group = "label", color = viridis(100),
                   cluster_rows = FALSE, cluster_cols = FALSE)
# Gene expression heatmap
dittoHeatmap(sce, top_markers, annot.by = "label", cluster_rows = TRUE)
```

# Marker gene detection (binomial test)

## Detecting marker genes

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
markers_binom <- findMarkers(sce, groups = sce$label, test.type = "binom",
                             pval.type = "all", direction = "up", lfc = 1)
markers_binom
head(markers_binom[["alpha"]])
```

## Top marker for each cell type

```{r, message = FALSE, warning = FALSE}
top_markers_binom <- lapply(markers_binom, function(x) rownames(x)[1]) %>%
	unlist() %>%
	enframe()
top_markers_binom
```

## Top marker gene tables {.tabset}

```{r, message = FALSE, warning = FALSE, results = 'asis'}
for (cell_type in levels(sce$label)) {
	cat("\n\n###", cell_type, "\n")
	markers_binom[[cell_type]] %>%
		as.data.frame() %>%
		head(n = 5) %>%
		t() %>%
		kable() %>%
		print()
}
```

## Top marker gene dot plot

```{r, message = FALSE, warning = FALSE}
top_markers_binom <- sapply(markers_binom, function(x) rownames(x)[1])
dittoDotPlot(sce, top_markers_binom, group.by = "label")
```

## Top marker gene heatmap

```{r, message = FALSE, warning = FALSE}
top_markers_binom <- sapply(markers_binom, function(x) rownames(x)[1])
dittoHeatmap(sce, top_markers_binom, annot.by = "label",
             scaled.to.max = TRUE, cluster_rows = FALSE)
```

# Session Info

```{r}
sessionInfo()
```
