---
title: "Analysis of mouse spermatogenic cells data (Hermann et al. 2018)"
author: "Michał Kabza"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: default
    highlight: pygments
    df_print: kable
    toc: true
    toc_depth: 2
    toc_float: true
---

# Preparing the environment

Load the required packages:
```{r, message = FALSE}
library(tidyverse)
library(glue)
library(knitr)
library(Matrix)
library(scRNAseq)
library(scater)
library(scran)
library(velociraptor)
library(dittoSeq)
library(viridis)
```

# Preparing the data

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Load the Hermann et al. (2018) mouse spermatogenic cells dataset
sce <- HermannSpermatogenesisData()
sce
# Cell metadata
colData(sce)
# Cell type summary
sce$celltype %>%
	table(useNA = "ifany") %>%
	enframe() %>%
	arrange(desc(value))
# Prepare the cell type labels
sce$cell_type <- sce$celltype %>%
  fct_explicit_na(na_level = "Other") %>%
	fct_lump_min(min = 400, other_level = "Other") %>%
	fct_relevel("Early Round spermatids", "Mid Round spermatids",
	            "Late Round spermatids", "Other") %>%
	fct_drop()
table(sce$cell_type)
```

# Basic analysis

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Set the RNG seed for reproducible results
set.seed(42)
# Data normalization
quick_clusters <- quickCluster(sce, assay.type = "spliced")
table(quick_clusters)
sce <- computeSumFactors(sce, clusters = quick_clusters, assay.type = "spliced")
sce <- logNormCounts(sce, assay.type = "spliced")
# Selecting highly variable genes
dec <- modelGeneVar(sce)
top_hvgs <- getTopHVGs(dec, fdr.threshold = 0.05)
length(top_hvgs)
# Principal component analysis
sce <- denoisePCA(sce, dec, subset.row = top_hvgs, min.rank = 5, max.rank = 50)
ncol(reducedDim(sce, "PCA"))
# Dimensionality reduction by UMAP
sce <- runUMAP(sce, dimred = "PCA")
```

# UMAP plots

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "cell_type", reduction.use = "UMAP", size = 0.5,
             legend.show = FALSE, do.label = TRUE,
             labels.highlight = FALSE, labels.size = 4)
dittoDimPlot(sce, "cell_type", reduction.use = "UMAP", size = 0.1,
             split.by = "cell_type", split.ncol = 2, legend.show = FALSE)
```

# RNA velocity analysis

## Steady-state mode {.tabset}

### scVelo

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Set the RNG seed for reproducible results
set.seed(42)
# Run scVelo
sce_velo_steady_state <- scvelo(sce,
                                mode = "steady_state",
                                assay.X = "spliced",
                                assay.spliced = "spliced",
                                assay.unspliced = "unspliced",
                                subset.row = top_hvgs,
                                use.dimred = "PCA")
sce_velo_steady_state
# Add RNA velocity information to the SingleCellExperiment object
sce$velocity_pseudotime_steady_state <- sce_velo_steady_state$velocity_pseudotime
sce$velocity_length_steady_state <- sce_velo_steady_state$velocity_length
sce$velocity_confidence_steady_state <- sce_velo_steady_state$velocity_confidence
# Project velocity vectors to the UMAP coordinates
embedded_steady_state <- embedVelocity(reducedDim(sce, "UMAP"),
                                       sce_velo_steady_state)
grid_df_steady_state <- gridVectors(reducedDim(sce, "UMAP"),
                                    embedded_steady_state, resolution = 30)
```

### RNA velocity vectors & pseudotime

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_pseudotime_steady_state",
             reduction.use = "UMAP", size = 0.5) +
	geom_segment(
		data = grid_df_steady_state,
		mapping = aes(x = start.1, y = start.2, xend = end.1, yend = end.2),
		arrow = arrow(length = unit(0.05, "inches"), type = "open")
	)
```

### RNA velocity vectors

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_pseudotime_steady_state",
             cells.use = rep(FALSE, ncol(sce)),
             reduction.use = "UMAP", size = 0.5) +
	geom_segment(
		data = grid_df_steady_state,
		mapping = aes(x = start.1, y = start.2, xend = end.1, yend = end.2),
		arrow = arrow(length = unit(0.05, "inches"), type = "open")
	)
```

### RNA velocity pseudotime

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_pseudotime_steady_state",
             reduction.use = "UMAP", size = 0.5)
```

### RNA velocity pseudotime by cell type

```{r, message = FALSE, warning = FALSE}
dittoPlot(sce, "velocity_pseudotime_steady_state", group.by = "cell_type",
          x.reorder = rev(seq(length(levels(sce$cell_type)))),
          plots = c("ridgeplot"), ridgeplot.lineweight = 0.5,
          legend.show = FALSE)
```

### RNA velocity length

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_length_steady_state",
             reduction.use = "UMAP", size = 0.5)
```

### RNA velocity confidence

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_confidence_steady_state",
             reduction.use = "UMAP", size = 0.5)
```

## Stochastic mode {.tabset}

### scVelo

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Set the RNG seed for reproducible results
set.seed(42)
# Run scVelo
sce_velo_stochastic <- scvelo(sce,
                              mode = "stochastic",
                              assay.X = "spliced",
                              assay.spliced = "spliced",
                              assay.unspliced = "unspliced",
                              subset.row = top_hvgs,
                              use.dimred = "PCA")
sce_velo_stochastic
# Add RNA velocity information to the SingleCellExperiment object
sce$velocity_pseudotime_stochastic <- sce_velo_stochastic$velocity_pseudotime
sce$velocity_length_stochastic <- sce_velo_stochastic$velocity_length
sce$velocity_confidence_stochastic <- sce_velo_stochastic$velocity_confidence
# Project velocity vectors to the UMAP coordinates
embedded_stochastic <- embedVelocity(reducedDim(sce, "UMAP"),
                                     sce_velo_stochastic)
grid_df_stochastic <- gridVectors(reducedDim(sce, "UMAP"),
                                  embedded_stochastic, resolution = 30)
```

### RNA velocity vectors & pseudotime

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_pseudotime_stochastic",
             reduction.use = "UMAP", size = 0.5) +
	geom_segment(
		data = grid_df_stochastic,
		mapping = aes(x = start.1, y = start.2, xend = end.1, yend = end.2),
		arrow = arrow(length = unit(0.05, "inches"), type = "open")
	)
```

### RNA velocity vectors

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_pseudotime_stochastic",
             cells.use = rep(FALSE, ncol(sce)),
             reduction.use = "UMAP", size = 0.5) +
	geom_segment(
		data = grid_df_stochastic,
		mapping = aes(x = start.1, y = start.2, xend = end.1, yend = end.2),
		arrow = arrow(length = unit(0.05, "inches"), type = "open")
	)
```

### RNA velocity pseudotime

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_pseudotime_stochastic",
             reduction.use = "UMAP", size = 0.5)
```

### RNA velocity pseudotime by cell type

```{r, message = FALSE, warning = FALSE}
dittoPlot(sce, "velocity_pseudotime_stochastic", group.by = "cell_type",
          x.reorder = rev(seq(length(levels(sce$cell_type)))),
          plots = c("ridgeplot"), ridgeplot.lineweight = 0.5,
          legend.show = FALSE)
```

### RNA velocity length

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_length_stochastic",
             reduction.use = "UMAP", size = 0.5)
```

### RNA velocity confidence

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_confidence_stochastic",
             reduction.use = "UMAP", size = 0.5)
```

## Dynamical mode {.tabset}

### scVelo

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Set the RNG seed for reproducible results
set.seed(42)
# Run scVelo
sce_velo_dynamical <- scvelo(sce,
                             mode = "dynamical",
                             assay.X = "spliced",
                             assay.spliced = "spliced",
                             assay.unspliced = "unspliced",
                             subset.row = top_hvgs,
                             use.dimred = "PCA")
sce_velo_dynamical
# Add RNA velocity information to the SingleCellExperiment object
sce$velocity_pseudotime_dynamical <- sce_velo_dynamical$velocity_pseudotime
sce$velocity_length_dynamical <- sce_velo_dynamical$velocity_length
sce$velocity_confidence_dynamical <- sce_velo_dynamical$velocity_confidence
# Project velocity vectors to the UMAP coordinates
embedded_dynamical <- embedVelocity(reducedDim(sce, "UMAP"),
                                    sce_velo_dynamical)
grid_df_dynamical <- gridVectors(reducedDim(sce, "UMAP"),
                                 embedded_dynamical, resolution = 30)
```

### RNA velocity vectors & pseudotime

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_pseudotime_dynamical",
             reduction.use = "UMAP", size = 0.5) +
	geom_segment(
		data = grid_df_dynamical,
		mapping = aes(x = start.1, y = start.2, xend = end.1, yend = end.2),
		arrow = arrow(length = unit(0.05, "inches"), type = "open")
	)
```

### RNA velocity vectors

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_pseudotime_dynamical",
             cells.use = rep(FALSE, ncol(sce)),
             reduction.use = "UMAP", size = 0.5) +
	geom_segment(
		data = grid_df_dynamical,
		mapping = aes(x = start.1, y = start.2, xend = end.1, yend = end.2),
		arrow = arrow(length = unit(0.05, "inches"), type = "open")
	)
```

### RNA velocity pseudotime

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_pseudotime_dynamical",
             reduction.use = "UMAP", size = 0.5)
```

### RNA velocity pseudotime by cell type

```{r, message = FALSE, warning = FALSE}
dittoPlot(sce, "velocity_pseudotime_dynamical", group.by = "cell_type",
          x.reorder = rev(seq(length(levels(sce$cell_type)))),
          plots = c("ridgeplot"), ridgeplot.lineweight = 0.5,
          legend.show = FALSE)
```

### RNA velocity length

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_length_dynamical",
             reduction.use = "UMAP", size = 0.5)
```

### RNA velocity confidence

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "velocity_confidence_dynamical",
             reduction.use = "UMAP", size = 0.5)
```

# Session Info

```{r}
sessionInfo()
```
