---
title: "Analysis of pbmc4k dataset (10X Genomics)"
author: "Michał Kabza"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: default
    highlight: pygments
    df_print: kable
    toc: true
    toc_depth: 2
    toc_float: true
---

# Preparing the environment

Load the required packages:
```{r, message = FALSE}
library(tidyverse)
library(glue)
library(Matrix)
library(TENxPBMCData)
library(scater)
library(scran)
library(SingleR)
library(celldex)
library(bluster)
library(dittoSeq)
library(Nebulosa)
```

# Preparing the data

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Load the pbmc4k peripheral blood mononuclear cell dataset
sce <- TENxPBMCData(dataset = "pbmc4k")
sce
# Cell metadata
head(colData(sce))
# Gene metadata
rowData(sce)
# Add column names to the SingleCellExperiment object
colnames(sce) <- sce$Barcode
rownames(colData(sce)) <- colnames(sce)
# Change row names to unique gene symbols
rownames(sce) <- uniquifyFeatureNames(rowData(sce)$ENSEMBL_ID, rowData(sce)$Symbol_TENx)
sce
```

# Basic analysis

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Set the RNG seed for reproducible results
set.seed(42)
# Data normalization
quick_clusters <- quickCluster(sce)
table(quick_clusters)
sce <- computeSumFactors(sce, clusters = quick_clusters)
sce <- logNormCounts(sce)
# Selecting highly variable genes
dec <- modelGeneVar(sce)
top_hvgs <- getTopHVGs(dec, fdr.threshold = 0.05)
length(top_hvgs)
# Principal component analysis
sce <- denoisePCA(sce, dec, subset.row = top_hvgs, min.rank = 5, max.rank = 50)
ncol(reducedDim(sce, "PCA"))
# Dimensionality reduction by UMAP
sce <- runUMAP(sce, dimred = "PCA")
```

# Cell type annotation

**Note**: Information about all available reference datasets can be found in the `r BiocStyle::Biocpkg("celldex")` package documentation

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Prepare the reference dataset
ref_se <- BlueprintEncodeData()
ref_se
# Annotate cell types using SingleR (main labels)
singler_results <- SingleR(sce, ref_se, labels = ref_se$label.main)
head(singler_results)
```

```{r, message = FALSE, warning = FALSE}
# SingleR predictions summary
table(singler_results$labels)
table(singler_results$pruned.labels)
sum(is.na(singler_results$pruned.labels))
# SingleR score heatmap
plotScoreHeatmap(singler_results)
# Add cell type information to the SingleCellExperiment object
sce$cell_type_blueprint_main <- singler_results$pruned.labels %>%
	fct_explicit_na(na_level = "Other") %>%
	fct_lump_min(min = 100, other_level = "Other") %>%
	fct_relevel("Other", after = Inf) %>%
	fct_drop()
# UMAP plots
dittoDimPlot(sce, "cell_type_blueprint_main", reduction.use = "UMAP", size = 0.5)
dittoDimPlot(sce, "cell_type_blueprint_main", reduction.use = "UMAP", size = 0.1,
             split.by = "cell_type_blueprint_main", split.ncol = 3)
```

# Cell clustering

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Set the RNG seed for reproducible results
set.seed(42)
# Detect cell clusters using a graph-based approach
colLabels(sce) <- clusterRows(reducedDim(sce, "PCA"),
                              NNGraphParam(k = 100, cluster.fun = "louvain"))
table(sce$label)
```

```{r, message = FALSE, warning = FALSE}
# UMAP plot
dittoDimPlot(sce, "label", reduction.use = "UMAP", size = 0.5)
# Cluster vs cell type barplots
dittoBarPlot(sce, "cell_type_blueprint_main", group.by = "label",
             scale = "percent")
dittoBarPlot(sce, "cell_type_blueprint_main", group.by = "label",
             scale = "count")
```

# Fine-grained cell type annotation

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Annotate cell types using SingleR (fine labels)
sce$cell_type_blueprint_fine <- SingleR(sce, ref_se, labels = ref_se$label.fine)$pruned.labels
table(sce$cell_type_blueprint_fine)
```

```{r, message = FALSE, warning = FALSE}
# Add cell type information to the SingleCellExperiment object
sce$cell_type_blueprint_fine <- sce$cell_type_blueprint_fine %>%
	fct_explicit_na(na_level = "Other") %>%
	fct_lump_min(min = 100, other_level = "Other") %>%
	fct_relevel("Other", after = Inf) %>%
	fct_drop()
# SingleR predictions summary
table(sce$cell_type_blueprint_fine)
# UMAP plot
dittoDimPlot(sce, "cell_type_blueprint_fine", reduction.use = "UMAP", size = 0.5)
# Cluster vs cell type barplots
dittoBarPlot(sce, "cell_type_blueprint_fine", group.by = "label",
             scale = "percent")
dittoBarPlot(sce[, colLabels(sce) %in% c(1, 4)],
             "cell_type_blueprint_fine", group.by = "label",
             scale = "percent")
```

# Using known marker genes

## Comparing monocyte clusters {.tabset}

```{r, message = FALSE, warning = FALSE}
# Identify marker genes differentiating the monocyte clusters
markers_monocyte <- findMarkers(sce, groups = sce$label, test.type = "wilcox",
                                pval.type = "any", direction = "up",
                                restrict = c(3, 5))
```

### Cluster 3 markers

```{r, message = FALSE, warning = FALSE}
head(as.data.frame(markers_monocyte[["3"]]), n = 10)
```

### Cluster 5 markers

```{r, message = FALSE, warning = FALSE}
head(as.data.frame(markers_monocyte[["5"]]), n = 10)
```

### CD14+ vs CD16+ monocytes scatter plot

```{r, message = FALSE, warning = FALSE}
dittoScatterPlot(sce[, colLabels(sce) %in% c(3, 5)],
                 x.var = "CD14", y.var = "FCGR3A", color.var = "label")
```

### CD14+ vs CD16+ monocytes hex scatter plot

```{r, message = FALSE, warning = FALSE}
dittoScatterHex(sce[, colLabels(sce) %in% c(3, 5)],
                x.var = "CD14", y.var = "FCGR3A",
                color.var = "label", max.density = 10)
```

## Comparing CD4+ T cell subpopulations {.tabset}

```{r, message = FALSE, warning = FALSE}
# Identify marker genes differentiating the CD4+ T cell subpopulations
sce$cell_type_cluster <- paste(sce$cell_type_blueprint_main, sce$label, sep = ".")
markers_cd4 <- findMarkers(sce, groups = sce$cell_type_cluster, test.type = "wilcox",
                           pval.type = "any", direction = "up",
                           restrict = c("CD4+ T-cells.1", "CD4+ T-cells.4"))
```

### Cluster 1 markers

```{r, message = FALSE, warning = FALSE}
head(as.data.frame(markers_cd4[["CD4+ T-cells.1"]]), n = 10)
```

### Cluster 4 markers

```{r, message = FALSE, warning = FALSE}
head(as.data.frame(markers_cd4[["CD4+ T-cells.4"]]), n = 10)
```

### Memory CD4+ T cell markers density plot

```{r, message = FALSE, warning = FALSE}
density_plots <- plot_density(sce, c("CD4", "IL7R", "S100A4"), size = 0.5,
                              joint = TRUE, combine = FALSE)
density_plots[[length(density_plots)]]
```

### Activated CD4+ T cell markers density plot

```{r, message = FALSE, warning = FALSE}
density_plots <- plot_density(sce, c("CD4", "IL7R", "CD69"), size = 0.5,
                              joint = TRUE, combine = FALSE)
density_plots[[length(density_plots)]]
```

### Naive CD4+ T cell markers density plot

```{r, message = FALSE, warning = FALSE}
density_plots <- plot_density(sce, c("CD4", "IL7R", "CCR7"), size = 0.5,
                              joint = TRUE, combine = FALSE)
density_plots[[length(density_plots)]]
```

# Session Info

```{r}
sessionInfo()
```
