---
title: "RNA velocity analysis and cell fate mapping using scVelo and Cellrank"
subtitle: "Data: mouse spermatogenic cells (Hermann et al. 2018)"
author: "Michał Kabza"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: default
    highlight: pygments
    df_print: kable
    toc: true
    toc_depth: 2
    toc_float: true
---

# Preparing the environment

Load the required packages:
```{r, message = FALSE}
library(tidyverse)
library(glue)
library(knitr)
library(Matrix)
library(scRNAseq)
library(scater)
library(scran)
library(dittoSeq)
library(reticulate)
library(zellkonverter)
```

Create a conda environment containing the required Python packages:
```{r, eval = FALSE}
if (!file.exists(miniconda_path())) {
	install_miniconda()
}
conda_install("r-cellrank", c("cellrank-krylov", "python-igraph"),
              python_version = "3.8", channel = c("conda-forge", "bioconda"))
```

Miniconda installation path:
```{r, message = FALSE, warning = FALSE}
miniconda_path()
```

List the available conda environments:
```{r, message = FALSE, warning = FALSE}
conda_list()
```

Use the created conda environment:
```{r, message = FALSE, warning = FALSE}
use_condaenv("r-cellrank", required = TRUE)
```

# Preparing the data

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Load the Hermann et al. (2018) mouse spermatogenic cells dataset
sce <- HermannSpermatogenesisData()
sce
# Fix row names
rownames(sce) <- str_remove(rownames(sce), ".\\d+$")
# Cell metadata
colData(sce)
# Cell type summary
sce$celltype %>%
  fct_explicit_na(na_level = "NA") %>%
	table() %>%
	enframe() %>%
	arrange(desc(value))
# Prepare the cell type labels
sce$cell_type <- sce$celltype %>%
  fct_explicit_na(na_level = "Other") %>%
  str_remove(" spermatids") %>%
	fct_lump_min(min = 400, other_level = "Other") %>%
	fct_relevel("Early Round", "Mid Round", "Late Round", "Other") %>%
	fct_drop()
table(sce$cell_type)
```

# Basic analysis

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
# Set the RNG seed for reproducible results
set.seed(42)
# Data normalization
quick_clusters <- quickCluster(sce, assay.type = "spliced")
table(quick_clusters)
sce <- computeSumFactors(sce, clusters = quick_clusters, assay.type = "spliced")
sce <- logNormCounts(sce, assay.type = "spliced")
# Selecting highly variable genes
dec <- modelGeneVar(sce)
top_hvgs <- getTopHVGs(dec, fdr.threshold = 0.05)
length(top_hvgs)
# Principal component analysis
sce <- denoisePCA(sce, dec, subset.row = top_hvgs, min.rank = 5, max.rank = 50)
ncol(reducedDim(sce, "PCA"))
# Dimensionality reduction by UMAP
sce <- runUMAP(sce, dimred = "PCA")
```

# UMAP plots

```{r, message = FALSE, warning = FALSE}
dittoDimPlot(sce, "cell_type", reduction.use = "UMAP", size = 0.5,
             legend.show = FALSE, do.label = TRUE,
             labels.highlight = FALSE, labels.size = 4)
dittoDimPlot(sce, "cell_type", reduction.use = "UMAP", size = 0.1,
             split.by = "cell_type", split.ncol = 2, legend.show = FALSE)
```

# Preparing the anndata object

```{r, message = FALSE, warning = FALSE}
sce_adata <- SingleCellExperiment(
  assays = list(
    counts = assay(sce, "spliced"),
    spliced = assay(sce, "spliced"),
    unspliced = assay(sce, "unspliced")
  ),
  colData = DataFrame(
    cell_type = sce$cell_type
  ),
  reducedDims = list(
    X_umap = reducedDim(sce, "UMAP")
  )
)
sce
adata <- SCE2AnnData(sce_adata)
adata
```

# RNA velocity analysis using scVelo

## Preparing the data

Load the required Python modules:
```{python}
import scanpy as sc
import scvelo as scv
import cellrank as cr
import matplotlib.pyplot as plt
scv.settings.verbosity = 0
cr.settings.verbosity = 0
```

Get the anndata object:
```{python}
adata = r.adata
```

Pie chart of spliced/unspliced proportions:
```{python}
scv.pl.proportions(adata, groupby='cell_type')
```

Preprocess the data:
```{python}
scv.pp.filter_and_normalize(adata, n_top_genes=2000, enforce=True)
scv.pp.moments(adata, n_pcs=30, n_neighbors=30)
```

## RNA velocity

RNA velocity analysis using the dynamical model:
```{python, results = 'hide'}
scv.tl.recover_dynamics(adata, n_jobs = 4)
scv.tl.velocity(adata, mode='dynamical')
scv.tl.velocity_graph(adata)
```

Stream plot of velocities on the UMAP embedding:
```{python}
scv.pl.velocity_embedding_stream(adata, basis='umap', color='cell_type')
```

## Speed and coherence

```{python}
scv.tl.velocity_confidence(adata)
scv.pl.scatter(adata, basis='umap', color='velocity_length',
               cmap='coolwarm', perc=[5, 95])
scv.pl.scatter(adata, basis='umap', color='velocity_confidence',
               cmap='coolwarm', perc=[5, 95])
```

## Velocity graph and pseudotime

```{python}
scv.tl.velocity_pseudotime(adata)
scv.pl.velocity_graph(adata, basis='umap', color='cell_type', threshold=0.1,
                      legend_loc='best')
scv.pl.scatter(adata, basis='umap', color='velocity_pseudotime',
               cmap='gnuplot')
```

## PAGA velocity graph

```{python}
scv.tl.paga(adata, groups='cell_type')
scv.pl.paga(adata, basis='umap', size=50, alpha=0.1, min_edge_width=2,
            node_size_scale=1.5, legend_loc='best')
```

## Differential velocity analysis

```{python}
scv.tl.rank_velocity_genes(adata, groupby='cell_type', min_corr=0.3)
diff_velo_df = scv.DataFrame(adata.uns['rank_velocity_genes']['names'])
```

```{r}
head(py$diff_velo_df)
```

```{python}
scv.pl.velocity(adata, diff_velo_df.iloc[0, :], color='cell_type', ncols = 1)
```

## Inspecting the anndata object

```{python}
adata
```

# Cell fate mapping using Cellrank

## Data processing using VelocityKernel

Calculating the transition matrix:
```{python, results = 'hide'}
vtk = cr.tl.kernels.VelocityKernel(adata)
vtk.compute_transition_matrix()
```

Plot the UMAP projection of the transition matrix:
```{python}
vtk.compute_projection(basis='umap')
scv.pl.velocity_embedding_stream(adata, basis='umap', color='cell_type',
                                 vkey='T_fwd')
```

## Macrostate analysis

Computing Schur decomposition:
```{python, results = 'hide'}
gpcaa = cr.tl.estimators.GPCCA(vtk)
gpcaa.compute_schur(n_components=10)
```

Top eigenvalues plot:
```{python}
gpcaa.plot_spectrum(real_only=True)
plt.show()
```

Computing macrostates:
```{python}
gpcaa.compute_macrostates(n_states=2, cluster_key='cell_type')
gpcaa.plot_macrostates()
```

Computing terminal states:
```{python}
gpcaa.compute_terminal_states()
gpcaa.plot_terminal_states()
```

## Absorption probabilities

Calculating absorption probabilities:
```{python, results = 'hide'}
gpcaa.compute_absorption_probabilities()
```

Plot the absorption probabilities:
```{python}
gpcaa.plot_absorption_probabilities(legend_loc='best')
gpcaa.plot_absorption_probabilities(same_plot=False, ncols=1)
```

## Inspecting the anndata object

```{python}
adata
```

## Saving the anndata object

```{python}
adata.uns['terminal_states_names'] = adata.obsm['to_terminal_states'].names
del adata.uns['coarse_fwd']
adata.write('rna_velocity_scvelo_hermann_2018.h5ad', compression='gzip')
```

# Getting back to R

Read the h5ad file:
```{r, message = FALSE, warning = FALSE}
sce_cellrank <- readH5AD("rna_velocity_scvelo_hermann_2018.h5ad")
sce_cellrank
```

Get terminal state absorption probabilities for each cell:
```{r, message = FALSE, warning = FALSE}
absorption_probabilities <- setNames(
  as.data.frame(reducedDim(sce_cellrank, "to_terminal_states")),
  metadata(sce_cellrank)$terminal_states_names
)
head(absorption_probabilities)
```

# Session Info

```{r}
sessionInfo()
```
