---
title: "Standard scran workflow"
description: "Analysis of the BaronPancreasData dataset using standard scran workflow"
author: "Michał Kabza"
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_depth: 1
---

```{r, include = FALSE}
library(BiocStyle)
```

# Preparing the environment

```{r, message = FALSE}
# Load the required packages
library(dplyr)
library(forcats)
library(tibble)
library(glue)
library(ggplot2)
library(pheatmap)
library(viridis)
library(clustree)
library(Matrix)
library(PCAtools)
library(SummarizedExperiment)
library(SingleCellExperiment)
library(scRNAseq)
library(scuttle)
library(bluster)
library(scater)
library(scran)
library(igraph)
library(BiocParallel)

# Check the default back-end for parallel evaluation
bpparam()
```

# Preparing the data

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
# Load the Baron et al. (2016) human pancreas dataset
sce <- BaronPancreasData('human')
sce
# UMI count per cell statistics
summary(colSums(counts(sce)))
# Expressed genes per cell statistics
summary(colSums(counts(sce) > 0))
# Cell type summary
colData(sce) %>%
	as.data.frame() %>%
	ggplot(mapping = aes(x = fct_rev(fct_infreq(label)))) +
	geom_bar() +
	labs(
		title = "Cell type summary",
		x = "Cell type",
		y = "Cell count"
	) +
	coord_flip() +
	theme_bw()
colData(sce)$label %>%
	table() %>%
	enframe() %>%
	arrange(desc(value))
# Remove cell types with fewer than 100 cells
cell_types_to_keep <- colData(sce)$label %>%
	table() %>%
	enframe() %>%
	filter(value >= 100) %>%
	pull(name)
sce <- sce[, colData(sce)$label %in% cell_types_to_keep]
sce
```

# Normalizing data

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
# Calculate per-cell size factors from the library sizes (i.e. total sum of counts per cell)
sce <- computeLibraryFactors(sce)
summary(sizeFactors(sce))
# scran offers a more advanced deconvolution strategy for size factor calculation
quick_clusters <- quickCluster(sce)
table(quick_clusters)
sce <- computeSumFactors(sce, clusters = quick_clusters)
summary(sizeFactors(sce))
# Calculate log-transformed normalized counts
sce <- logNormCounts(sce)
sce
```

# Selecting highly variable genes

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
# Model gene variance
dec <- modelGeneVar(sce)
dec
dec %>%
	as.data.frame() %>%
	ggplot(mapping = aes(x = mean, y = total)) +
	geom_point() +
	geom_smooth() +
	labs(
		title = "Modelling gene variance",
		x = "Mean log-expression",
		y = "Total variance"
	) +
	theme_bw()
# Select the top 2000 variable genes
top_hvgs <- getTopHVGs(dec, n = 2000)
length(top_hvgs)
# Select the top 10% of variable genes
top_hvgs <- getTopHVGs(dec, prop = 0.1)
length(top_hvgs)
length(top_hvgs) / nrow(dec)
# Overall number of variable genes
length(getTopHVGs(dec, prop = 1))
# Select all variable genes with FDR below 5%
top_hvgs <- getTopHVGs(dec, fdr.threshold = 0.05)
length(top_hvgs)
```

# Principal component analysis

**Option 1**
```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
# Set the RNG seed for reproducible results
set.seed(42)
# Run the PCA
sce <- runPCA(sce, ncomponents = 50, subset_row = top_hvgs)
sce
ncol(reducedDim(sce, "PCA"))
# Use scree plot to choose the number of PCs
percent_var <- attr(reducedDim(sce, "PCA"), "percentVar")
pca_elbow <- findElbowPoint(percent_var)
pca_elbow
ggplot(mapping = aes(x = seq_along(percent_var), y = percent_var)) +
geom_point() +
geom_vline(xintercept = pca_elbow, col = "blue") +
labs(
	title = "Scree plot",
	x = "PC",
	y = "Variance explained (%)"
) +
theme_bw()
# Keep only selected PCs
reducedDim(sce, "PCA") <- reducedDim(sce, "PCA")[, 1:pca_elbow, drop = FALSE]
```

**Option 2**
```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
# Set the RNG seed for reproducible results
set.seed(42)
# Run the PCA
sce <- runPCA(sce, ncomponents = 50, subset_row = top_hvgs)
ncol(reducedDim(sce, "PCA"))
# Use clusters to choose the number of PCs
pca_stats <- getClusteredPCs(reducedDim(sce, "PCA"), k = 10)
npcs <- metadata(pca_stats)$chosen
npcs
pca_stats %>%
	as.data.frame() %>%
	ggplot(mapping = aes(x = n.pcs, y = n.clusters)) +
	geom_point() +
	labs(
		title = "getClusteredPCs results",
		x = "Number of PCs",
		y = "Number of clusters"
	) +
	theme_bw()
# Keep only selected PCs
reducedDim(sce, "PCA") <- reducedDim(sce, "PCA")[, 1:npcs, drop = FALSE]
```

**Option 3**
```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
# Set the RNG seed for reproducible results
set.seed(42)
# Run the PCA and remove principal components corresponding to technical noise
sce <- denoisePCA(sce, dec, subset.row = top_hvgs, min.rank = 5, max.rank = 50)
ncol(reducedDim(sce, "PCA"))
```

# Dimensionality reduction

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
# Set the RNG seed for reproducible results
set.seed(42)
# Run UMAP
sce <- runUMAP(sce, dimred = "PCA")
plotUMAP(sce, colour_by = "label", text_by = "label", text_size = 3) +
	theme(legend.position = "none")
```

# Cell clustering

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
# Set the RNG seed for reproducible results
set.seed(42)
# Build the nearest-neighbor graph
snn_graph <- buildSNNGraph(sce, use.dimred = "PCA", k = 10)
```

**Detect clusters using short random walks (Walktrap clustering)**

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
set.seed(42)
colData(sce)$cluster_walktrap <- cluster_walktrap(snn_graph)$membership
table(colData(sce)$cluster_walktrap)
plotUMAP(sce, colour_by = "cluster_walktrap", text_by = "cluster_walktrap", text_size = 3) +
	theme(legend.position = "none")
cluster_table <- table(fct_infreq(colData(sce)$label),
                       colData(sce)$cluster_walktrap)
pheatmap(log10(cluster_table + 10),
         main = "Cell type vs Walktrap clusters",
         color = viridis(100),
         cluster_rows = FALSE, cluster_cols = TRUE,
         treeheight_row = 0, treeheight_col = 0)
cluster_modularity <- pairwiseModularity(snn_graph, colData(sce)$cluster_walktrap, as.ratio = TRUE)
pheatmap(log10(cluster_modularity + 1),
         main = "Walktrap cluster modularity",
         color = viridis(100),
         cluster_rows = FALSE, cluster_cols = FALSE)
```

**Detect clusters using multi-level optimization of modularity (Louvain clustering)**

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
set.seed(42)
colData(sce)$cluster_louvain <- cluster_louvain(snn_graph)$membership
table(colData(sce)$cluster_louvain)
plotUMAP(sce, colour_by = "cluster_louvain", text_by = "cluster_louvain", text_size = 3) +
	theme(legend.position = "none")
cluster_table <- table(fct_infreq(colData(sce)$label),
                       colData(sce)$cluster_louvain)
pheatmap(log10(cluster_table + 10),
         main = "Cell type vs Louvain clusters",
         color = viridis(100),
         cluster_rows = FALSE, cluster_cols = TRUE,
         treeheight_row = 0, treeheight_col = 0)
cluster_modularity <- pairwiseModularity(snn_graph, colData(sce)$cluster_louvain, as.ratio = TRUE)
pheatmap(log10(cluster_modularity + 1),
         main = "Louvain cluster modularity",
         color = viridis(100),
         cluster_rows = FALSE, cluster_cols = FALSE)
```

**Using different values of k**

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
set.seed(42)
for (k in c(5, 10, 20, 50, 100)) {
	snn_graph <- buildSNNGraph(sce, use.dimred = "PCA", k = k)
	colData(sce)[glue("k{k}")] <- cluster_louvain(snn_graph)$membership
}
clustree(sce, prefix = "k") +
	guides(edge_alpha = FALSE, edge_colour = FALSE)
plotUMAP(sce, colour_by = "k100", text_by = "k100", text_size = 3) +
	theme(legend.position = "none") +
	labs(title = "Louvain clustering (k = 100)")
```

# Marker gene detection

**Detecting marker genes using t-test**

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
markers <- findMarkers(sce, groups = sce$label, test.type = "t",
                       pval.type = "all", direction = "up", lfc = 1)
markers
head(markers[["alpha"]])
# Get top marker for each cell type
top_markers <- lapply(markers, function(x) {rownames(x)[1]}) %>%
	unlist() %>%
	enframe()
top_markers
# Top marker expression heatmap 
plotGroupedHeatmap(sce, features = top_markers$value, group = "label",
                   color = viridis(100), cluster_rows = FALSE, cluster_cols = FALSE)
# Show marker gene expression on a UMAP plot
plotUMAP(sce, colour_by = "GCG")
plotUMAP(sce, colour_by = "INS")
```

**Detecting marker genes using Wilcoxon rank sum test**

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}

markers <- findMarkers(sce, groups = sce$label, test.type = "wilcox",
                       pval.type = "all", direction = "up")
head(markers[["alpha"]])
# Get top marker for each cell type
top_markers <- lapply(markers, function(x) {rownames(x)[1]}) %>%
	unlist() %>%
	enframe()
top_markers
```

**Detecting marker genes using binomial test**

```{r, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE, dpi = 200}
markers <- findMarkers(sce, groups = sce$label, test.type = "binom",
                       pval.type = "all", direction = "up")
head(markers[["alpha"]])
# Get top marker for each cell type
top_markers <- lapply(markers, function(x) {rownames(x)[1]}) %>%
	unlist() %>%
	enframe()
top_markers
```

# Session Info

```{r}
sessionInfo()
```

